---
title: "Patterns in Paediatric ASD Screening: An Exploratory Data Analysis"
author: "Ifeanyichukwu Herbertson"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

Autistic Spectrum Disorder (ASD) is a neurodevelopmental condition 
characterised by differences in social communication, behaviour, and 
sensory processing. This analysis explores a children's ASD screening 
dataset sourced from the UCI Machine Learning Repository, collected by 
Fadi Thabtah using the AQ-10-Child screening tool. The dataset contains 
292 records across 21 variables, and this report investigates the 
distribution of ASD classifications, the relationship between screening 
scores and ASD outcome, demographic associations, and the discriminatory 
power of individual screening questions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Paste your library() calls here
library(foreign)
library(tidyverse)
library(janitor)
```

# Data Loading and Cleaning

```{r cleaning}
df <- read.arff("Autism-Child-Data.arff")

df <- df %>% 
  rename(ASD = `Class/ASD`) %>% 
  clean_names()

df <- df %>% 
  mutate(across(a1_score:a10_score, ~as.integer(as.character(.))))

df <- df %>% select(-age_desc)

df <- df %>% filter(!is.na(age))
```

# Results

## 1. Distribution of ASD Classifications

```{r plot1, echo=FALSE}
ggplot(df, aes(x = asd, fill = asd)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("NO" = "#4C9BE8", "YES" = "#E8734C")) +
  labs(title = "Distribution of ASD Classification",
       x = "ASD Classification", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

The dataset contains 149 children classified as non-ASD (NO) and 139 
classified as ASD (YES), representing a near-equal 52/48 split. This 
balance is unusual for clinical screening data and likely reflects 
deliberate curation for research purposes rather than real-world 
prevalence rates.

## 2. Screening Score by ASD Classification

```{r plot2, echo=FALSE}
ggplot(df, aes(x = asd, y = result, fill = asd)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("NO" = "#4C9BE8", "YES" = "#E8734C")) +
  labs(title = "Screening Score by ASD Classification",
       x = "ASD Classification",
       y = "Screening Score (result)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Children classified with ASD show consistently higher screening scores, 
with a median of approximately 8, compared to a median of around 5 for 
non-ASD children. The ASD group also shows much less variability, 
suggesting that a threshold score of around 7 may be a strong indicator 
of ASD classification in this dataset.

## 3. Demographic Patterns

```{r plot3, echo=FALSE}
# Plot 3a: ASD Classification by Gender
ggplot(df, aes(x = gender, fill = asd)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("NO" = "#4C9BE8", "YES" = "#E8734C")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "ASD Classification by Gender",
       x = "Gender", y = "Proportion", fill = "ASD") +
  theme_minimal()

# Plot 3b: ASD Classification by Jaundice History
ggplot(df, aes(x = jundice, fill = asd)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("NO" = "#4C9BE8", "YES" = "#E8734C")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "ASD Classification by Jaundice at Birth",
       x = "Born with Jaundice", y = "Proportion", fill = "ASD") +
  theme_minimal()
```

Gender shows no meaningful difference in ASD classification rates, with 
both males and females showing an approximately equal split between YES 
and NO outcomes. Similarly, jaundice at birth does not appear to be a 
strong differentiator, with children born with and without jaundice 
showing comparable ASD classification proportions.

## 4. AQ-10 Question Scores

```{r plot4, echo=FALSE}
df %>%
  select(a1_score:a10_score, asd) %>%
  pivot_longer(cols = a1_score:a10_score,
               names_to = "question",
               values_to = "score") %>%
  mutate(question = fct_relevel(question, 
                                "a1_score", "a2_score", "a3_score", 
                                "a4_score", "a5_score", "a6_score",
                                "a7_score", "a8_score", "a9_score", 
                                "a10_score")) %>%
  group_by(question, asd) %>%
  summarise(avg_score = mean(score), .groups = "drop") %>%
  ggplot(aes(x = question, y = avg_score, fill = asd)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("NO" = "#4C9BE8", "YES" = "#E8734C")) +
  labs(title = "Average Score per AQ-10 Question by ASD Classification",
       x = "Question", y = "Average Score (0-1)", fill = "ASD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Across almost all ten AQ-10 questions, children classified with ASD 
scored 1 at a higher average rate than non-ASD children. Questions A4, 
A8, A9, and A10 show the largest gaps between the two groups, suggesting 
these items are the most discriminating between ASD and non-ASD children.

# Summary

This exploratory analysis of the ASD screening dataset reveals several 
clear patterns. Screening scores are strongly separated between ASD and 
non-ASD groups, with ASD-classified children scoring consistently higher 
and within a tighter range. Demographic variables such as gender and 
jaundice history show little association with ASD classification in this 
dataset. At the question level, items A4, A8, A9, and A10 appear to be 
the most useful discriminators between the two groups. Overall, the 
AQ-10-Child screening tool shows strong potential as a classification 
aid, with screening score alone appearing to be a reliable indicator of 
ASD outcome in this sample.

# Dataset Limitations

Several limitations should be considered when interpreting these findings. 
First, the dataset contains only 288 records after cleaning, which is a 
relatively small sample for drawing broad conclusions about ASD screening 
patterns across populations. Second, the near-equal split between ASD and 
non-ASD cases (52/48) is unlikely to reflect real-world prevalence, where 
ASD affects an estimated 1-2% of the general population, suggesting the 
dataset was deliberately balanced for research and classification purposes. 
Third, the dataset draws heavily from certain countries and ethnic groups, 
limiting how generalisable the findings are to a global population. Finally, 
this analysis is exploratory in nature — the patterns identified here describe 
associations in the data and should not be interpreted as clinical diagnostic 
criteria.